---
title: "Transaction coding"
author: "Joseph Crispell"
date: "`r format(Sys.time(), '%d-%m-%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
params:
  transactions_file: "dummy_transactions.csv"
  transaction_coding_file: "dummy_transaction_coding_dictionary.csv"
---

```{r setup, include=FALSE}
# Load required packages
library(rmarkdown) # building this report
library(DT) # interactive tables
library(plotly) # interactive graphics

# Set default code chunk options
knitr::opts_chunk$set(echo = FALSE)

# Note data folder
data_folder <- file.path("..", "data")

# Note key parameters
date_column <- "Date"
date_format <- "%Y-%m-%d"
description_column <- "Description"
in_column <- "In"
out_column <- "Out"

# Note input files
transactions_file <- file.path(data_folder, params$transactions_file)
transaction_coding_file <- file.path(
  data_folder,
  params$transaction_coding_file
)

# Note building report
building_report <- TRUE

# Source plotting functions
source("plotting_functions.R")
```

```{r data_processing, include = FALSE}
# Run the data analysis - note the variables set in setup chunk will
# be carried through
source("process_transactions.R")
```

# Introduction

The following report summarises monthly bank transactions based upon a key word(s) classification table. Different types of transactions are recognised based upon key words that are found.

The key word classification table  used for the current report is shown here with key word(s) for each type in the `Patterns` column separated by semi colons (`;`):
```{r types_table}
datatable(transaction_types, rownames = FALSE)
```

Bank transactions were provided for the from `r format(transactions_dates[1], "%B %Y")` to `r format(transactions_dates[2], "%B %Y")` drawing information from the following columns:

- ``r date_column`` - column containing transaction date
- ``r description_column`` - payment description column
- ``r in_column`` - amount of money paid in
- ``r out_column`` - amount of money paid out

With the following additional columns add based on the type classifications:

- `Type` - transaction type classification
- `Pattern` - pattern that matched for type classification

```{r transactions}
datatable(transactions[, c(
  date_column, in_column,
  out_column, description_column,
  "Type", "Pattern"
)],
rownames = FALSE
)
```

# Unclassified transactions

Based on the above type classifications using key words, the following transaction descriptions didn't contain any key words:

```{r unclassified_counts}
datatable(summary_of_unclassifieds, rownames = FALSE)
```

# Average monthly earnings and costs {.tabset}

Without considering transaction type the average monthly earnings, costs, and savings are the following:

- Earnings: `£`r format(round(average_in, 0), big.mark=",")``
- Costs: `£`r format(round(average_out, 0), big.mark=",")``
- Savings: `£`r format(round(average_savings, 0), big.mark=",")``

Here is how the above values vary by month:

## Static

```{r plot_monthly_totals_static}
plot_values_by_month(
  totals_by_month[-nrow(totals_by_month), ],
  title = "Total money in and out"
)
```

## Interactive

```{r plot_monthly_totals_interactive}
plotly_values_by_month(
  totals_by_month[-nrow(totals_by_month), ],
  title = "Total money in and out"
)
```

## Data

```{r monthly_totals_data}
datatable(totals_by_month[-nrow(totals_by_month), ], rownames = FALSE)
```

# Monthly earnings and costs by type

## {.tabset}

Considering the transaction type we can break down the monthly earnings, costs, and savings on average across the months considered:

### Static

```{r type_average_static}
# Plot the values
plot_values_bar_chart(
  values = average_by_type$difference,
  names = average_by_type$Type,
  main = "Average value by type",
  cex.names = 0.5
)
```

### Interactive
```{r type_average_interactive}
# Create the bar chart
plotly_values_bar_chart(
  values = average_by_type$difference,
  names = average_by_type$Type,
  title = "Average value by type"
)
```

### Data
```{r type_average_data}
datatable(average_by_type, rownames = FALSE)
```

## {.tabset}
And by month:

### Static
```{r type_by_month_static}
# Plot each type by month
plot_values_by_month(
  type_difference_by_month,
  title = "Money in/out by type",
  legend_cex = 0.45
)
```

### Interactive
```{r type_by_month_interactive}
# Plot each type by month
plotly_values_by_month(
  type_difference_by_month,
  title = "Money in/out by type"
)
```

### Data
```{r type_by_month_data}
# Add months as column at start rather than just row names
type_difference_by_month <- cbind(
  rownames(type_difference_by_month),
  type_difference_by_month
)
colnames(type_difference_by_month)[1] <- "month"

# Show data
datatable(
  type_difference_by_month,
  extensions = "FixedColumns",
  options = list(
    dom = "t",
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 1)
  ),
  rownames = FALSE
)
```
